Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2024-06-05 manuscript.Rmd
a bunch of updates, now its ready to submit

---
title: 'Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
author:
  - Kaleonani K.C. Hurley:
      institute:
        - himb
      equal_contributor: yes
      correspondence: yes
      email: "hurleyk@hawaii.edu"
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
  - Kerry Reardon:
      institute: noaa
  - Derek J. Skillings:
      institute: 
        - uncg
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology, University of Hawai&#x02BB;i, Mānoa
  - noaa: National Oceanic and Atmospheric Administration
  - uncg: 'University of North Carolina, Greensboro'
  - natgeo: National Geographic Pristine Seas
abstract:
  Coral reef communities are influenced by various environmental and anthropogenic factors, yet those factors have not been consistently resolved. Furthermore, due to logistical difficulties, investigations of the role of depth (particularly mesophotic coral ecosystems) in influencing coral reef community structure have been taxonomically limited. Here, we examined the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of shallow brachyuran crabs across the ~2,500 km span of the Hawaiian Archipelago. In addition, we compared shallow crab assemblages with those sampled on a depth transect from 12--90 m on O&#x02BB;ahu. We deployed Autonomous Reef Monitoring Structures (ARMS) on shallow reefs (12--16 m) across the latitudinal gradient of the Hawaiian Archipelago to compare with a depth gradient (12, 30, 60, and 90 m) on the south shore of O&#x02BB;ahu. We found a total of 2,284 individual crabs representing 103 (morpho)species in 60 genera and 22 families. Shallow crab assemblages were significantly different between the Main and Northwest Hawaiian Islands and variation in community structure was significantly associated with six biotic, environmental, and anthropogenic factors (cumulative human impacts, chlorophyll-A, sea-surface temperature, site depth, island slope, and potential larval connectivity). In depth comparisons, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among all shallow samples across the complete span of the archipelago. We show that 90 m of depth is a more potent determinant of brachyuran assemblage composition than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.<!--Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems. Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to ~170 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago.-->
bibliography: references/citations.json
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--toc-depth=2"
      - "--filter=resources/pandoc-crossref"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--reference-doc=resources/ref.docx"
    keep_md: true
tblPrefix: table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>

@@

```{r init, include=FALSE}
source("crabs.R")
if (!exists("crabs_done")) {
  crabs_done <<- FALSE
}
if (!crabs_done) {
  cc <<- setup_crabs()
  crabs_done <<- TRUE
}

@@

The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019]. Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongterm2003]. In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zone or mesophotic reefs intensifies. To date, the majority of such efforts have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018]. However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010]. Here, we investigate the factors influencing the community structure of brachyuran crabs across a suite of anthropogenic and environmental gradients using the Hawaiian Archipelago as a natural laboratory.

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean and is the most isolated island chain on the planet. Typically regarded as two regions, the main Hawaiian Islands (MHI) include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) comprise the uninhabited islands, atolls, and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau. The NWHI are wholly enclosed by the Papahānaumokuākea Marine National Monument, one of the largest marine protected areas in the world. The islands are subject to considerable anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which are widely believed to influence coral reef communities. These gradients include differences in temperature, coral cover, human population, and anthropogenic impact, among others. Studies examining community structure across broad environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. We then sampled brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011] from these structures to examine drivers of community composition across these gradients. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.

@@

One aspect of coral reef community structure that is thought to be important but remains under-investigated is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021]. Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow reef brachyuran crab assemblages vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (ii) Do shallow reef brachyuran crab assemblages differ between the Main and Northwestern Hawaiian Islands? (iii) What environmental and anthropogenic factors drive crab assemblage structure in shallow coral reefs across the Hawaiian Archipelago? (iv) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

@@

Brachyuran crab assemblages were sampled from benthic coral reef habitat in the Hawaiian Archipelago using Autonomous Reef Monitoring Structures (ARMS). ARMS are standardized collection devices designed to mimic coral reef structural complexity to attract settlement of reef cryptofauna [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. They are constructed from 10 gray type 1 PVC plates (23 cm $\times$ 23 cm) stacked in an alternating series of open and semi-enclosed layers ([@fig:fig_arms]). The stack is fastened to a larger base plate (35 cm $\times$ 45 cm) which can be attached to the benthos upon deployment [@brainardAutonomousReefMonitoring2009; @hurleyAssessmentShallowMesophotic2016; @knowltonCoralReefBiodiversity2010; @zimmermanArtificialReefMatrix2004].

![An autonomous reef monitoring structure (ARMS) unit deployed on a reef in the Red Sea (photo credit: Michael Berumen, used with permission)](data/images/arms_example.png){#fig:fig_arms}

@@

A total of 89 ARMS units were deployed over the course of six years (2010--2016). Units were deployed by NOAA divers during Pacific Reef Assessment and Monitoring Program (RAMP) cruises in shallow coral reef habitat (12--16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]). To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in 12--90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].

@@

```{r arms-table,results="asis", dependson="init"}
arms_schedule <- cc$crabs_untransformed %>%
  sample_tibble() %>%
  separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
  select(sample,everything()) %>%
  unite("range",deployment_year,recovery_year,sep="--") %>%
  mutate(
    island = case_when(
      str_remove_all(hawaiian_name,'[^a-zA-Z ]') != island ~ as.character(str_glue("{hawaiian_name} ({island})")),
      TRUE ~ as.character(hawaiian_name)
    ),
    island = case_when(
      island == "O‘ahu" ~ str_glue("{hawaiian_name} ({shallow_deep})"),
      TRUE ~ island
    ),
    island = fct_reorder(island,-lat)
  ) %>%
  count(island,range) %>%
  pivot_wider(names_from="range",values_from="n")  %>%
  select(island,order(colnames(.))) %>%
  mutate(across(-island,~replace_na(.x,0)))
print_table(arms_schedule,"arms_table","ARMS unit deployment schedule (number of units)",accuracy=1) 
```

@@

ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details]. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified, and preserved in 95% ethanol. Identifications were made to species level where possible, otherwise crabs were classified into morphospecies represented by suites of shared morphological characters.<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->

@@

To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and Simpson diversity (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species). We used nonparametric Kruskal-Wallis ranked sum tests to compare diversity between shallow and mesophotic samples and between island groups (MHI vs. NWHI). Island group comparisons were performed both including and excluding mesophotic samples. <!--To explore environmental influences on alpha diversity, we used a linear mixed-effect model with Simpson or richness as response variables, the environmental variables listed below as fixed effects, and island nested within island group (MHI vs NWHI) as random effects. Models were generated using the `lmer` function from the `lme4` package [@batesFittingLinearMixedEffects2015]. -->

We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables to crab community variation across the islands. We first used a stepwise forward selection procedure (`ordiR2step`) to find the variables that best explained community dissimilarity. Then, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated. Model significance was assessed using permutation tests with `r num(PERM,1)` permutations with the function `anova`. Continuous variables were scaled to unit variance.

@@

Potential drivers of shallow crab community structure were assembled from a previous study examining environmental effects on genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers (PO.DAAC, OB.DAAC) ([@tbl:env_table]). For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental material. Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).

@@

Overall mean species richness for the complete dataset was `r num(all_alpha$richness,0.1)` $\pm$ `r num(all_alpha$richness_sd,0.1)` species per unit (Simpson diversity: `r num(all_alpha$simpson,0.01)` $\pm$ `r num(all_alpha$simpson_sd,0.01)`). Overall mean species richness for shallow ARMS \< 30 m was `r num(shallow_alpha$richness,0.1)` $\pm$ `r num(shallow_alpha$richness_sd,0.1)` species per unit (Simpson diversity: `r num(shallow_alpha$simpson,0.01)` $\pm$ `r num(shallow_alpha$simpson_sd,0.01)`). Overall mean species richness for mesophotic ARMS $\geq$ 30 m was `r num(deep_alpha$richness,0.1)` $\pm$ `r num(deep_alpha$richness_sd,0.1)` species per unit (Simpson diversity: `r num(deep_alpha$simpson,0.01)` $\pm$ `r num(deep_alpha$simpson_sd,0.01)`).

@@

For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}a]; `r k_test(all_alpha$k_r_island)`), but Simpson diversity was significantly different ([@fig:fig_alpha[]{}c]; `r k_test(all_alpha$k_s_island,plain=TRUE)`). Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but Simpson diversity was ([@fig:fig_alpha[]{}d]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`). When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).

<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r w_test(all_alpha$w_s_island,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`). Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).  -->

>>>>

---
title: 'Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
author:
  - Kaleonani K.C. Hurley:
      institute:
        - himb
      equal_contributor: yes
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
      correspondence: yes
      email: "mhoban@hawaii.edu"
  - Kerry Reardon:
      institute: noaa
  - Derek J. Skillings:
      institute: 
        - uncg
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology, University of Hawai&#x02BB;i, Mānoa
  - noaa: National Oceanic and Atmospheric Administration
  - uncg: 'University of North Carolina, Greensboro'
  - natgeo: National Geographic Pristine Seas
abstract:
  An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change. Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs. Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (8--17 m) across the ~2,500 km span of the Hawaiian Archipelago. We compare these assemblages with those sampled along a depth gradient (12--90 m) on O&#x02BB;ahu. We found 2,284 individuals representing 103 (morpho)species from 22 families. Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval connectivity, and depth. However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.
bibliography: references/citations.json
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--toc-depth=2"
      - "--filter=resources/pandoc-crossref"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--reference-doc=resources/ref.docx"
    keep_md: false
tblPrefix: table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>

@@

```{r init, include=FALSE}
source("crabs.R")

if (!exists("crabs_done")) {
  crabs_done <<- FALSE
}
if (!crabs_done) {
  cc <<- setup_crabs()
  crabs_done <<- TRUE
}

@@

Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongterm2003]. In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zones or mesophotic reefs intensifies. While the structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019]. One aspect of coral reef community structure that is thought to be important but remains under-investigated is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].

The Hawaiian Archipelago, stretching across ~2,500 km in the North Pacific Ocean, provides a natural laboratory to examine drivers of metazoan community structure. This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---the Papahānaumokuākea Marine National Monument. The island chain is subject to considerable anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reef communities [@selkoeMapHumanImpacts2009]. Nearshore drop-offs along the island chain also allow access to deeper reef habitat, enabling the examination of depth influences on community structure.

To date, the majority of studies that have examined community structure across the Hawaiian archipelago and along a depth gradient have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018]. However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010]. Furthermore, studies on these taxa have also been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].

Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. We targeted brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011], recovered from these structures to examine drivers of community composition across the Hawaiian Archipelago. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.

@@

Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow reef brachyuran crab assemblages vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (ii) Do shallow reef brachyuran crab assemblages differ between the Main and Northwestern Hawaiian Islands? (iii) What environmental and anthropogenic factors drive crab assemblage structure in shallow coral reefs across the Hawaiian Archipelago? (iv) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

@@

Brachyuran crab assemblages were sampled from benthic coral reef habitat in the Hawaiian Archipelago using Autonomous Reef Monitoring Structures (ARMS) collected by NOAA's Ecosystem Science Division as part of its Pacific Reef Assessment and Monitoring Program. ARMS are standardized collection devices designed to mimic coral reef structural complexity to attract settlement of reef cryptofauna [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. They are constructed from 10 gray type 1 PVC plates (23 cm $\times$ 23 cm) stacked in an alternating series of open and semi-enclosed layers ([@fig:fig_arms]). The stack is fastened to a larger base plate (35 cm $\times$ 45 cm) which can be attached to the benthos upon deployment [@brainardAutonomousReefMonitoring2009; @hurleyAssessmentShallowMesophotic2016; @knowltonCoralReefBiodiversity2010; @zimmermanArtificialReefMatrix2004].

![Autonomous reef monitoring structures (ARMS) deployed in the Hawaiian Archipelago.](data/images/arms_example.png){#fig:fig_arms}

@@

A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (8--17 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]). To explore the effects of depth, we also present data from an earlier study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in 12--90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]). Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].

@@

```{r arms-table,results="asis", dependson="init"}
arms_schedule <- cc$crabs_untransformed %>%
  sample_tibble() %>%
  # separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
  # select(sample,everything()) %>%
  unite("range",deployment_year,recovery_year,sep="--") %>%
  mutate(
    island = case_when(
      str_remove_all(hawaiian_name,'[^a-zA-Z ]') != island ~ as.character(str_glue("{hawaiian_name} ({island})")),
      TRUE ~ as.character(hawaiian_name)
    ),
    island = case_when(
      island == "O‘ahu" ~ str_glue("{hawaiian_name} ({shallow_deep})"),
      TRUE ~ island
    ),
    island = fct_reorder(island,-lat)
  ) %>%
  count(island,range) %>%
  pivot_wider(names_from="range",values_from="n")  %>%
  select(island,order(colnames(.))) %>%
  mutate(across(-island,~replace_na(.x,0)))
print_table(arms_schedule,"arms_table","Recovered ARMS unit soak times across sites (number of units)",accuracy=1) 
```

@@

ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details]. All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, and preserved in 95% ethanol. Identifications were made to species level where possible, otherwise crabs were classified into morphospecies represented by suites of shared morphological characters according to expert advice and Poore & Ahyong [-@pooreMarineDecapodCrustacea2023].<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->

@@

To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and effective Simpson diversity (effective Simpson = $\frac{1}{\sum_{i=1}^{R}p_i^2}$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species). The effective Simpson index (the inverse of the "classic" Simpson entropy), which has units of species, is easier to interpret in an ecological context than an entropic non-linear index [@jostEntropyDiversity2006; @tuomistoUpdatedConsumerGuide2012]. We used nonparametric Kruskal-Wallis ranked sum tests to compare diversity between shallow and mesophotic samples and between island groups (MHI vs. NWHI). Island group comparisons were performed both including and excluding mesophotic samples.

We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:env_table]) to crab community variation across the islands, conditioning the recovery year as a random effect. From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated. Model significance was assessed using permutation tests with `r num(PERM,1)` permutations with the function `anova`. Continuous variables were scaled to unit variance.

@@

Potential drivers of shallow crab community structure were assembled from a previous study examining environmental effects on genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers (PO.DAAC, OB.DAAC) ([@tbl:env_table; @tbl:arms_metadata_table]). Sea-surface temparature (SST) and chlorophyll (Chl-A) values were calculated by averaging monthly means for the duration of the soaking period at each sampling site. For details on satellite data preparation and data sources, see Supplemental material. Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).

@@

Overall mean species richness for the complete dataset was `r num(all_alpha$richness,0.1)` $\pm$ `r num(all_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(all_alpha$simpson,0.01)` $\pm$ `r num(all_alpha$simpson_sd,0.01)`). Overall mean species richness for shallow ARMS \< 30 m was `r num(shallow_alpha$richness,0.1)` $\pm$ `r num(shallow_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(shallow_alpha$simpson,0.01)` $\pm$ `r num(shallow_alpha$simpson_sd,0.01)`). Overall mean species richness for mesophotic ARMS $\geq$ 30 m was `r num(deep_alpha$richness,0.1)` $\pm$ `r num(deep_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(deep_alpha$simpson,0.01)` $\pm$ `r num(deep_alpha$simpson_sd,0.01)`).

@@

For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}a]; `r k_test(all_alpha$k_r_island)`), but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}c]; `r k_test(all_alpha$k_s_island,plain=TRUE)`). Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but effective Simpson diversity was ([@fig:fig_alpha[]{}d]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`). When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).

<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r w_test(all_alpha$w_s_island,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; effective Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`). Effective Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).  -->

<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r t_test(all_alpha$t_s_island,plain=TRUE)`). When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; effective Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`). Effective Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).  -->
